---
title: "Two-sex time-varying kinship model specified by age"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{
Two-sex time-varying kinship model specified by age
}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Set up code chunk options
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE, 
                      warning = FALSE, 
                      fig.align = 'center',
                      fig.width = 8,
                      fig.height = 6,
                      dpi = 300)
# Prevent scientific notation (useful for the rate calculation)
options(scipen = 999999)
```

<style>
/* Make TOC appear on the right side */
.tocify {
  position: fixed;
  top: 50px;
  right: 0;
  width: 240px;
  height: calc(100% - 50px);
  overflow-y: auto;
  margin-left: 0;
  padding-left: 0;
}

/* Adjust main content to make room for right-side TOC */
.main-container {
  max-width: calc(100% - 260px);
  margin-left: 10px;
  margin-right: 250px;
}

/* Limit TOC to only show 2 levels */
.tocify-subheader .tocify-subheader {
  display: none;
}

/* Style for learning objectives box */
.learning-objectives {
  background-color: #e0f7fa;
  border-left: 5px solid #00acc1;
  padding: 15px;
  margin-bottom: 20px;
  border-radius: 3px;
}
</style>

<div class="learning-objectives">
<strong>Learning Objectives</strong>: In this vignette, you will learn how to implement a kinship model that combines both two-sex and time-varying approaches. You will understand how to incorporate sex-specific demographic rates that change over time, analyze the effects of demographic transition on kinship structures by sex, and explore approximation methods when data is limited.
</div>

# Introduction {#introduction}

Kinship networks are shaped by two fundamental forces: demographic differences between males and females, and historical changes in demographic rates over time. While we have explored these dimensions separately in previous vignettes, this vignette combines them into a more comprehensive framework: a **two-sex time-varying kinship model**.

This integrated approach offers several key advantages:

- Accounts for sex differences in mortality and fertility patterns
- Incorporates historical demographic changes from the demographic transition
- Enables comparison between period and cohort perspectives
- Provides more realistic estimates of kinship structures by sex
- Allows for the analysis of sex ratios across time periods

In this vignette, we will implement a two-sex time-varying model using the `kin2sex` function with `time_invariant = FALSE`. This allows us to analyze how kinship networks evolve differently for males and females across changing demographic conditions.

## Package Installation {#preparation}

If you haven't already installed the required packages from the previous vignettes, here's what you'll need:

```{r installs, eval=FALSE}
# Install basic data analysis packages
install.packages("dplyr")     # Data manipulation
install.packages("tidyr")     # Data tidying
install.packages("ggplot2")   # Data visualization
install.packages("readr")     # Data import
install.packages("knitr")     # Document generation
install.packages("data.table")# Efficient data handling
install.packages("Matrix")    # Matrix operations

# Install DemoKin
# DemoKin is available on CRAN (https://cran.r-project.org/web/packages/DemoKin/index.html), 
# but we'll use the development version on GitHub (https://github.com/IvanWilli/DemoKin):
install.packages("remotes")
remotes::install_github("IvanWilli/DemoKin")
```

# Setting Up the Analysis Environment {#load-packages}

Let's load the necessary packages for our analysis:

```{r libraries, warning=F, message=FALSE}
rm(list = ls())
library(dplyr)    # For data manipulation
library(tidyr)    # For restructuring data
library(ggplot2)  # For visualization
library(readr)    # For reading data
library(knitr)    # For document generation
library(DemoKin)  # For kinship analysis
```

# Two-Sex Time-Varying Kinship Models {#model-overview}

## Model Structure and Components

The combined two-sex time-varying kinship model expands on previous models by requiring:

1. **Sex-specific mortality rates over time**: How survival differs between males and females across historical periods
2. **Sex-specific fertility rates over time**: How fertility patterns differ between males and females across historical periods
3. **Sex ratio at birth**: The proportion of births that are female
4. **Sex of the focal individual**: Whether we're analyzing male or female kinship networks

This model allows us to account for both sex differences and temporal changes simultaneously, providing a more realistic representation of kinship dynamics.

# Data Preparation {#model-input-2sex-time-varying}

For this vignette, we'll use Swedish demographic data from the `DemoKin` package. However, since the package only includes female fertility and mortality data over time, we'll create synthetic male rates for illustration purposes.

In a real-world application, you would ideally use actual male rates. Here, we'll create male rates by applying transformations to the female rates:

```{r}
# Extract dimensions of the data
years <- ncol(swe_px)
ages <- nrow(swe_px)

# Use female rates directly from the package
swe_surv_f_matrix <- swe_px
swe_fert_f_matrix <- swe_asfr

# Create synthetic male rates
# Male survival: Lower than female (raising to power 1.5 reduces values)
swe_surv_m_matrix <- swe_px ^ 1.5 # artificial perturbation for this example

# Male fertility: Shifted to slightly older ages and slightly higher
swe_fert_m_matrix <- rbind(matrix(0, 5, years),  
                          swe_asfr[-((ages-4):ages),]) * 1.05 # artificial perturbation for this example
```

Let's examine the resulting rates for a specific year (1900) to verify they follow expected patterns:

```{r}
bind_rows(
  data.frame(age = 0:100, sex = "Female", component = "Fertility rate", value = swe_fert_f_matrix[,"1900"]),
  data.frame(age = 0:100, sex = "Male", component = "Fertility rate", value = swe_fert_m_matrix[,"1900"]),
  data.frame(age = 0:100, sex = "Female", component = "Survival probability", value = swe_surv_f_matrix[,"1900"]),
  data.frame(age = 0:100, sex = "Male", component = "Survival probability", value = swe_surv_m_matrix[,"1900"])) %>% 
  ggplot(aes(age, value, col = sex)) +
  geom_line() +
  theme_bw() +
  facet_wrap(~component, scales = "free") +
  labs(
    title = "Sex-specific demographic rates in Sweden (1900)",
    x = "Age",
    y = "Rate",
    color = "Sex"
  )
```

**Interpretation**: The plot confirms our synthetic rates follow expected patterns:

- **Fertility**: Male fertility is shifted to slightly older ages compared to female fertility
- **Survival**: Male survival probabilities are lower than female survival at all ages
- **Both rates** show characteristic age patterns: fertility concentrated in reproductive ages, and survival declining with age

These sex differences will shape kinship networks in important ways, especially when combined with historical demographic changes.

# Running the Two-Sex Time-Varying Model {#run-model-2sex-time-varying}

Now, let's implement the two-sex time-varying kinship model using the `kin2sex` function with `time_invariant = FALSE`:

```{r}
kin_out_time_variant <- kin2sex(
  pf = swe_surv_f_matrix,       # Female survival matrix (age x year)
  pm = swe_surv_m_matrix,       # Male survival matrix (age x year)
  ff = swe_fert_f_matrix,       # Female fertility matrix (age x year)
  fm = swe_fert_m_matrix,       # Male fertility matrix (age x year)
  sex_focal = "f",              # Focal individual is female
  time_invariant = FALSE,       # Use time-varying model
  birth_female = .5,            # Sex ratio at birth (50% female)
  output_cohort = 1900          # Focus on the 1900 birth cohort
)
```

The resulting output provides detailed information on the kinship network for the 1900 female birth cohort, with relatives classified by both age and sex.

# Comparing Time-Variant and Time-Invariant Models {#kin-availability-2sex-time-varying}

To understand the impact of incorporating historical demographic change, let's compare the time-varying model with a time-invariant model that uses only 1900 rates:

```{r, message=FALSE, warning=FALSE}
# Run a time-invariant model using only 1900 rates
kin_out_time_invariant <- kin2sex(
  pf = swe_surv_f_matrix[,"1900"],    # Female survival (1900)
  pm = swe_surv_m_matrix[,"1900"],    # Male survival (1900)
  ff = swe_fert_f_matrix[,"1900"],    # Female fertility (1900) 
  fm = swe_fert_m_matrix[,"1900"],    # Male fertility (1900)
  sex_focal = "f",                    # Focal individual is female
  birth_female = .5                   # Sex ratio at birth (50% female)
)

# Combine and plot the results
kin_out_time_variant$kin_summary %>%
  filter(cohort == 1900) %>% 
  mutate(type = "variant") %>%
  bind_rows(kin_out_time_invariant$kin_summary %>% mutate(type = "invariant")) %>% 
  # Combine siblings and aunts/uncles for simplicity
  mutate(kin = case_when(kin %in% c("ys", "os") ~ "s",
                         kin %in% c("ya", "oa") ~ "a",
                         TRUE ~ kin)) %>%
  # Select key relationships
  filter(kin %in% c("d", "m", "gm", "ggm", "s", "a")) %>%
  # Group by relationship type, age, sex, and model type
  group_by(type, kin, age_focal, sex_kin) %>%
  summarise(count = sum(count_living)) %>%
  # Create plot
  ggplot(aes(age_focal, count, linetype = type)) +
  geom_line() + 
  theme_bw() +
  facet_grid(cols = vars(kin), rows = vars(sex_kin), scales = "free") +
  labs(
    title = "Time-variant vs. time-invariant kinship models by sex",
    subtitle = "Female 1900 birth cohort, Sweden",
    x = "Age of focal individual",
    y = "Number of living relatives",
    linetype = "Model type"
  )
```

**Interpretation**: This comparison reveals important differences between time-variant and time-invariant models:

- **Male relatives**: The differences are especially pronounced for male kin (top row), where mortality improvements over time led to greater availability than predicted by the time-invariant model
- **Ascending generations**: For mothers, grandmothers, and great-grandmothers, the time-varying model shows higher kin availability at older ages, reflecting mortality improvements not captured by the time-invariant model
- **Descendants**: For daughters, the differences are smaller but still present, particularly at later ages

The time-varying model captures the demographic transition that occurred over the 20th century, including declining fertility and mortality rates. This produces a more accurate representation of kinship dynamics for historical cohorts.

# Approximation Methods for Limited Data {#approximations}

In practice, demographic data is often limited, particularly for male fertility rates which can be difficult to obtain. Caswell [-@caswell_formal_2022] introduced two approximation methods to estimate two-sex kinship networks when male demographic rates are unavailable:

1. **Androgynous approximation**: Assumes equal fertility and survival for males and females
2. **GKP factors**: Applies multipliers to one-sex kin counts based on theoretical considerations

Let's evaluate these approximations using French data from the `DemoKin` package:

## Androgynous Approximation {#androgynous}

The androgynous approximation uses female rates for both sexes. Let's compare it to the full two-sex model:

```{r, message=FALSE, warning=FALSE}
# Full two-sex model
kin_out <- kin2sex(fra_surv_f, fra_surv_m, fra_fert_f, fra_fert_m, 
                  sex_focal = "f", birth_female = .5)

# Androgynous approximation
kin_out_androgynous <- kin2sex(fra_surv_f, fra_surv_f, fra_fert_f, fra_fert_f, 
                              sex_focal = "f", birth_female = .5)

# Compare the results
bind_rows(
  kin_out$kin_summary %>% mutate(type = "full"),
  kin_out_androgynous$kin_summary %>% mutate(type = "androgynous")) %>%
  group_by(kin, age_focal, sex_kin, type) %>%
  summarise(count = sum(count_living)) %>%
  ggplot(aes(age_focal, count, linetype = type)) +
  geom_line() +
  theme_bw() +
  theme(legend.position = "bottom", axis.text.x = element_blank()) +
  facet_grid(row = vars(sex_kin), col = vars(kin), scales = "free") +
  labs(
    title = "Androgynous approximation vs. full two-sex model",
    subtitle = "France, 2012",
    x = "Age of focal individual",
    y = "Number of living relatives",
    linetype = "Model type"
  )
```

**Interpretation**: The androgynous approximation performs well for most kin types, particularly for female relatives. However, it shows noticeable discrepancies for male grandparents and great-grandparents, where ignoring male-specific mortality leads to overestimation.

## GKP Factors Approximation {#gkp}

The GKP factors approach applies theoretical multipliers to one-sex kin counts. Let's implement and evaluate this approach:

```{r, message=FALSE, warning=FALSE}
# One-sex model
kin_out_1sex <- kin(fra_surv_f, fra_fert_f, birth_female = .5)

# Apply GKP factors
kin_out_GKP <- kin_out_1sex$kin_summary %>%
  mutate(count_living = case_when(
    kin == "m" ~ count_living * 2,          # Parents: multiply by 2
    kin == "gm" ~ count_living * 4,         # Grandparents: multiply by 4
    kin == "ggm" ~ count_living * 8,        # Great-grandparents: multiply by 8
    kin == "d" ~ count_living * 2,          # Children: multiply by 2
    kin == "gd" ~ count_living * 4,         # Grandchildren: multiply by 4
    kin == "ggd" ~ count_living * 4,        # Great-grandchildren: multiply by 4
    kin == "oa" ~ count_living * 4,         # Older aunts/uncles: multiply by 4
    kin == "ya" ~ count_living * 4,         # Younger aunts/uncles: multiply by 4
    kin == "os" ~ count_living * 2,         # Older siblings: multiply by 2
    kin == "ys" ~ count_living * 2,         # Younger siblings: multiply by 2
    kin == "coa" ~ count_living * 8,        # Cousins (older): multiply by 8
    kin == "cya" ~ count_living * 8,        # Cousins (younger): multiply by 8
    kin == "nos" ~ count_living * 4,        # Nieces/nephews (older): multiply by 4
    kin == "nys" ~ count_living * 4         # Nieces/nephews (younger): multiply by 4
  ))

# Compare approaches at selected ages
bind_rows(
  kin_out$kin_summary %>% mutate(type = "full"),
  kin_out_androgynous$kin_summary %>% mutate(type = "androgynous"),
  kin_out_GKP %>% mutate(type = "gkp")) %>%
  # Combine siblings, aunts/uncles, cousins, and nieces/nephews
  mutate(kin = case_when(
    kin %in% c("ys", "os") ~ "s",           # All siblings
    kin %in% c("ya", "oa") ~ "a",           # All aunts/uncles
    kin %in% c("coa", "cya") ~ "c",         # All cousins
    kin %in% c("nys", "nos") ~ "n",         # All nieces/nephews
    TRUE ~ kin)) %>%
  # Select specific ages for comparison
  filter(age_focal %in% c(5, 15, 30, 60, 80)) %>%
  # Sum across sex for total kin counts
  group_by(kin, age_focal, type) %>%
  summarise(count = sum(count_living)) %>%
  # Create bar chart
  ggplot(aes(type, count)) +
  geom_bar(aes(fill = type), stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90), legend.position = "bottom") +
  facet_grid(col = vars(kin), row = vars(age_focal), scales = "free") +
  labs(
    title = "Comparison of approximation methods with full two-sex model",
    subtitle = "France, 2012, at selected ages",
    x = "Approximation method",
    y = "Total number of living relatives",
    fill = "Method"
  )
```

**Interpretation**: This comparison shows that both approximation methods produce reasonable estimates for most kin types:

- **GKP factors**: Perform well for most kin types but may overestimate some relationships (especially more distant ones)
- **Androgynous**: Generally provides good approximations, particularly for close kin, but less accurate for distant male relatives
- **Age effects**: The accuracy of approximations varies by Focal's age, with greater discrepancies at older ages

These approximations offer practical alternatives when sex-specific data is limited, though they should be used with caution and their limitations understood.

# Incorporating Causes of Death {#prevalance-2sex-time-varying}

The `kin2sex` function also allows for the analysis of kinship networks by cause of death, providing insights into how different mortality causes affect kin availability. Let's implement a simple cause-of-death model with two competing causes.

We'll assume the following:
- For females, the risk of the first cause is half the risk of the second cause for ages above 50
- For males, the risk of the first cause is one-third the risk of the second cause for ages above 50
- Both sexes have equal risk of both causes before age 50

```{r}
# Create matrices of relative risks by cause, sex, and age
Hf <- matrix(c(.5, 1), nrow = 2, ncol = length(fra_surv_f))  # Female risk factors
Hm <- matrix(c(.33, 1), nrow = 2, ncol = length(fra_surv_f)) # Male risk factors

# Set equal risks for ages below 50
Hf[,1:50] <- Hm[,1:50] <- 1
```

Now we'll run the two-sex model with cause of death information:

```{r}
kin_out_cod_invariant <- kin2sex(
  pf = fra_surv_f,           # Female survival
  pm = fra_surv_m,           # Male survival
  ff = fra_fert_f,           # Female fertility
  fm = fra_fert_m,           # Male fertility
  Hf = Hf,                   # Female cause-specific risk factors
  Hm = Hm,                   # Male cause-specific risk factors
  time_invariant = TRUE      # Using time-invariant model for simplicity
)
```

Let's examine the structure of the output:

```{r}
head(kin_out_cod_invariant)
```

The output now includes additional columns for each cause of death. Let's visualize the distribution of parental deaths by cause, sex, and age for a 30-year-old Focal:

```{r}
kin_out_cod_invariant %>% 
  filter(kin == "m", age_focal == 30) %>% 
  summarise(deadcause1 = sum(deadcause1),
            deadcause2 = sum(deadcause2), .by = c(age_kin, sex_kin)) %>%
  pivot_longer(deadcause1:deadcause2) %>% 
  ggplot(aes(age_kin, value, col = sex_kin, linetype = name)) +
  geom_line() + 
  labs(
    title = "Distribution of parental deaths by cause, sex, and age",
    subtitle = "For a 30-year-old focal individual",
    x = "Age of parent at death",
    y = "Expected number of parental deaths",
    color = "Sex of parent",
    linetype = "Cause of death"
  ) +
  theme_bw()
```

**Interpretation**: This visualization shows the distribution of parental deaths by cause:

- **Timing**: In this simplified example, all parental deaths occur after age 50
- **Sex differences**: Male parents (fathers) show higher death counts at earlier ages, reflecting their higher mortality
- **Cause differences**: The relative importance of different causes varies by sex, with males showing a different distribution than females

This approach can be extended to include more causes of death and to incorporate time-varying cause-specific mortality, though this would require more complex data inputs.

# Conclusion

In this vignette, we've explored how to implement two-sex time-varying kinship models using the `DemoKin` package. These models provide a more comprehensive framework for understanding kinship dynamics by incorporating both sex differences and historical demographic change.

Key insights include:

1. Time-varying models capture the effects of demographic transition on kinship networks
2. Sex-specific models reveal important differences in the availability of male versus female relatives
3. Approximation methods offer practical alternatives when data is limited
4. Cause-of-death extensions provide insights into how different mortality causes shape kinship structures

These advanced models have numerous applications in demography, sociology, and public health, including:
- Understanding historical changes in support networks
- Analyzing sex differences in care provision and receipt
- Projecting future kinship structures under different demographic scenarios
- Examining how specific causes of death affect family structures

Future research might combine these approaches with multi-state models to create even more comprehensive frameworks for kinship analysis.

# References

